---
title: "Linear mixed-effects models on all variables"
format:
  html:
    toc: true
    code-fold: true
    code-overflow: wrap
    code-tools: true
---

```{r setup, include=F}
knitr::opts_chunk$set(echo = T, message = F, warning = F, eval = F)
```


```{r libs, eval = T}
library(dplyr)
library(lmerTest)
library(lme4)
library(here)
library(broom)
library(MultiAssayExperiment)

here::i_am("helper-lmm.qmd")
source("src/summary_lmm_v2.R")
```

This script is to run linear mixed models on all data points. The output is saved under `out/lmm_continuous.Rdata` or `out/lmm_factor.Rdata` and is loaded in for the final figures.

# Load data

Normalized data are loaded to run trajectories. To avoid any duplicates for variables (e.g., methylation scores computed in two tissues and assays), assay and variable name are concatenated.

```{r getdata, eval = F}
# load in data
load("data/data_normalized.Rdata")

# extract all data in long format plus basic anno
df <- as.data.frame(longForm(data, colData = c("age_at_consent",
                                                 "interventionId",
                                                 "subjectId",
                                                 "visitId", 
                                                 "time",
                                                 'compliance',
                                                 'smoking_py'))) |> 
  dplyr::filter(!is.na(value))

# Filter variables: (see t-tests.qmd for how this was generated:)
load(here("src/vars.Rdata"))

# remove non-S and any non M0-M6
df <- df |>
  dplyr::inner_join(vars, by = c('rowname' = 'x',
                                 'assay' = 'assay')) |> 
  dplyr::select(-c(label, assay2)) |>
  dplyr::filter(interventionId=='S' & !visitId %in% c("M12", "M18")) |>
  dplyr::mutate(compliance = factor(compliance, levels = c("lower compliance", "higher compliance"))) |> 
  dplyr::mutate(rowname = paste0(assay, "_", rowname)) 

# remove full data
rm(data);gc()

data  <- df
variables = unique(data$rowname)
```

# Linear model: factor

The linear model is run on each variable in the data, using visitId as a factor.

```{r, eval = F}
out_lmm <- summary_lmm(data,
                       variables = variables,
                       timeType = 'factor',
                       outName = 'out_lmm_factor.Rdata')
```


# Linear model: continuous

The linear model is run on each variable in the data, using time as a continuous variable (not used in main manuscript).

```{r lmm.cont, eval = F}
out_lmm <- summary_lmm(data,
                       variables = variables,
                       timeType = 'continuous',
                       outName = 'out_lmm_continuous_15Apr2024.Rdata')
```

```{r save.as.script, eval = F}
knitr::purl("helper-lmm.qmd",
            output = 'helper-lmm.R')

# To make it more efficient, the script can be exported into an R file and run from terminal/command line
```

# Write Extended Data Table (Factor)

```{r, eval = F}
load(here("out/out_lmm_factor.Rdata"))
source(here("src/renameVarsLME.R"))
out_lmm<- renameVarsLME(out_lmm)
```

# Write supplementary tables

Now we write the lmm to the extended data tables (xlsx):

```{r, eval=F}
for (x in c("Basic model with packyears", "Interaction model with packyears", "Higher compliance only with packyears")){
  file = case_when(x == 'Basic model with packyears' ~ "3",
                   x == 'Interaction model with packyears' ~ '5', 
                   x == 'Higher compliance only with packyears' ~ '7')
  
  writexl::write_xlsx(as.data.frame(out_lmm[[x]]),
                      path = paste0("out/Supplementary-Table-",
                                    file,
                                    ".xlsx")
    )
}
```

See also extended data table pages ([ST3](st3.qmd), [ST5](st5.qmd), [ST7](st7.qmd)) and data portal.
